Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-28 LICS21/antitone.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\end{theorem}

@@

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

@@

Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
\begin{align*}
T(\omega) & := 
\begin{cases}
\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
\infty & \hbox{otherwise}
\end{cases} 
\\
Y_k(\omega) & := 
\begin{cases}
4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
0 & \hbox{otherwise}
\end{cases} 
\end{align*}
\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
\end{example}

\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
%Recall that $(f(M_n))_{n \in \omega}$ is a supermartingale.
We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \in \omega}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \in \omega}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

\begin{proof}
If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(t_n)_{n \in \omega}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(y_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.

@@

The sequence $(y_n)_{n \in \omega}$ is then defined recursively by
\begin{align*}
    y_0 = 2, \quad
    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
\end{align*}

@@

The assumption that $(\mathcal F_n)_{n \in \omega}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

>>>>

\begin{definition}
We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\end{theorem}

@@

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \geq 0}$ actually takes (or at least if $(Y_n)_{n \geq 0}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

@@

Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
\begin{align*}
T(\omega) & := 
\begin{cases}
\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
\infty & \hbox{otherwise}
\end{cases} 
\\
Y_k(\omega) & := 
\begin{cases}
4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
0 & \hbox{otherwise}
\end{cases} 
\end{align*}
\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
Plainly, $T$ is a stopping time, and $(Y_n)_{n \geq 0}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \geq 0}$ where $\calF_n = \calF$ for all $n$. 
In this case, $(Y_n)_{n \geq 0}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \geq 0}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \geq 0}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
\end{example}

\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
%Recall that $(f(M_n))_{n \geq 0}$ is a supermartingale.
We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \geq 0}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \geq 0}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

\begin{proof}
If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.

@@

The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
\begin{align*}
    y_0 = 2, \quad
    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
\end{align*}

@@

The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

### 2020-12-28 LICS21/supermartingales.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
\end{definition}

@@

Fix a probability space $(\Omega, \calF, \mathbb{P})$. 
(In subsequent applications to AST analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)

\begin{definition}\rm
\begin{enumerate}
\item A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
\item Let $\epsilon > 0$. A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{enumerate}
\end{definition}

\begin{remark}
\changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
\end{remark}

@@

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

@@

Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \in \omega}$.
Recall the $\sigma$-algebra (consisting of measurable subsets ``prior to $T$'')
\[
\calF_T := \set{A \in \calF \mid \forall i \in \omega \, . \, A \cap \set{T \leq i} \in \calF_i}
\]
and if $T \leq T'$, then $\calF_{T} \subseteq \calF_{T'}$.

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

@@

%\emph{Notation.}
\changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}

Our first result is the following theorem.
\begin{theorem}[Rankability] 
\label{thm:rankable and strict rankable}
If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \in \omega}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$. %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
\end{theorem}

@@

%Thus the random variable $T_n(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the $(n+1)$-occurrence of a $\tY$-redex (in the reduction sequence $M_0(s), M_1(s), \ldots$).
%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$.
We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
\begin{restatable}{lemma}{TnBounded}
\label{lem:TnBounded}
$(T_n)_{n \in\omega}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.
\end{restatable}

%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \in \omega}$. 
%Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.

The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
Moreover 
\begin{restatable}{lemma}{TMtYStoppingTime}
\label{lem:TMtY is a stopping time}
$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \in \omega}$.
\end{restatable}

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For 1, since $(T_n)_{n \in\omega}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \in \omega}$ (\Cref{lem:TnBounded}),
and $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \in \omega}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

%For 2, we have that $(f(M_n))_{n \in \omega}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement 2 follows immediately from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
\end{proof}

>>>>

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.
\end{definition}

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \geq 0}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n \geq 0}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
\lo{It is confusing (strictly speaking, incorrect) to say that $(M_n)_{n \geq 0}$ is a reduction sequence from $M$. Formally $M_n$ is the random variable $M_n: s \mapsto \pi_0 (\red^n(M, s))$.}

@@

\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
(In subsequent applications to termination analysis, $\Omega = S$, and $\mathbb{P} = \mu$.)
Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \in \omega$.}

\begin{definition}\rm
\begin{enumerate}
\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
\item Let $\epsilon > 0$. A stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(s) := 1$ if $s \in A$, and $0$ otherwise.}
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{enumerate}
\end{definition}

\begin{remark}
\changed[lo]{Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
\end{remark}

@@

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

@@

Let $T$ and $T'$ be stopping times adapted to $(\calF_n)_{n \geq 0}$.
Recall the $\sigma$-algebra (consisting of measurable subsets ``prior to $T$'')
\[
\calF_T := \set{A \in \calF \mid \forall i \in \omega \, . \, A \cap \set{T \leq i} \in \calF_i}
\]
and if $T \leq T'$, then $\calF_{T} \subseteq \calF_{T'}$.

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n \geq 0}$ be a supermartingale, and $(T_n)_{n \geq 0}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \geq 0}$, then $(X_{T_n})_{n \geq 0}$ is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n \geq 0}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

@@

%\emph{Notation.}
\changed[lo]{For the rest of this section, fix the probability space $(S, \calF, \mu)$, and a closed SPCF term $M$.
For $n \in \omega$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}

Our first result is the following theorem.
\begin{theorem}[Rankability] 
\label{thm:rankable and strict rankable}
If a closed SPCF term $M$ is rankable (respectively, strictly rankable) by $f$ satisfying $f(N) = 0$ iff $N$ is a value, then $(f(M_n))_{n \geq 0}$ is a supermartingale (respectively, ranking supermartingale with stopping time $T_M: s \mapsto \min \set{n \mid f(M_n(s)) = 0}$) adapted to the filtration $(\mathcal{F}_n)_{n \geq 0}$. %where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
\end{theorem}

@@

%Thus the random variable $T_n(s)$ is the least $k \geq 0$ such that $M_k(s)$ is either a value or the $(n+1)$-occurrence of a $\tY$-redex (in the reduction sequence $M_0(s), M_1(s), \ldots$).
%Notice that each $T_n$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$.
We first state a useful property about r.v.~$T_0, T_1, T_2, \cdots$.
\begin{restatable}{lemma}{TnBounded}
\label{lem:TnBounded}
$(T_n)_{n \geq 0}$ is an increasing sequence of stopping times adapted to $(\calF_n)_{n \geq 0}$, and each $T_i$ is bounded.
\end{restatable}

%The random variable $T_M$ is a stopping time w.r.t.~filtration $(\calF_n)_{n \geq 0}$. 
%Since $f(M_n(s)) = 0$ iff $M_n(s)$ is a value, $T_M$ is the \emph{runtime of $M$}.

The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
Moreover 
\begin{restatable}{lemma}{TMtYStoppingTime}
\label{lem:TMtY is a stopping time}
$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \geq 0}$.
\end{restatable}

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is bounded and adapted to $(\calF_n)_{n \geq 0}$ (\Cref{lem:TnBounded}),
and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \geq 0}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement 2 follows immediately from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
\end{proof}

### 2020-12-28 akr-thots-completeness.tex
Global replace: _{n \in \omega} -> _{n \geq 0}

Take $N \in \mathit{Rch}(M)$.
Assume $N$ is not a value.
Define $\rho_N : \mathbb{N} \to \nnReal$ by setting
\[
\rho_N(n) := \mu(\{s \mid \pi_0(\red^n(N, s)) \textrm{~non-value}\})
\]
I.e.~$\rho_N(n)$ is the probability that $N$ is not yet terminated after $n$ more reduction steps.
\lo{E.g.~writing $I^{m+1} := I^m \, I$ where $I^1 := I$, we have $\rho_{I^m}(n) = 1$ if $n < m - 1$, and 0 otherwise.}
Let $|\rho_N| := \sum_{n \in \omega} \rho_N(n)$.

@@

\lo{Define r.v.~$Z_n(s) := \widehat{\rho}_N$ where $\red^n(M, s) = (N, {-})$. 
Then $(Z_n)_{n \in \omega}$ is a (generalised) supermartingale.
I am not sure if we need this.}

>>>>

Take $N \in \mathit{Rch}(M)$.
Assume $N$ is not a value.
Define $\rho_N : \mathbb{N} \to \nnReal$ by setting
\[
\rho_N(n) := \mu(\{s \mid \pi_0(\red^n(N, s)) \textrm{~non-value}\})
\]
I.e.~$\rho_N(n)$ is the probability that $N$ is not yet terminated after $n$ more reduction steps.
\lo{E.g.~writing $I^{m+1} := I^m \, I$ where $I^1 := I$, we have $\rho_{I^m}(n) = 1$ if $n < m - 1$, and 0 otherwise.}
Let $|\rho_N| := \sum_{n \geq 0} \rho_N(n)$.

@@

\lo{Define r.v.~$Z_n(s) := \widehat{\rho}_N$ where $\red^n(M, s) = (N, {-})$. 
Then $(Z_n)_{n \geq 0}$ is a (generalised) supermartingale.
I am not sure if we need this.}
